Parity games in pushdown systems with pebbles

### 2023-06-11 reachability/bibliography/intro.tex
remains from pushdown parameter article added in

\documentclass[times,envcountsame]{llncs}
\usepackage{a4wide}
\usepackage{times}
\usepackage{amsmath,amssymb,eufrak,theorem}%mnsymbol}
\usepackage{enumerate}
\usepackage{complexity}
\usepackage{gastex}

\usepackage{xspace,tabularx,multicol}

%\usepackage[thmmarks,standard,thref]{ntheorem}

\iffalse

\newtheorem{theorem}{{\bf Theorem}}[section]
\newtheorem{corollary}[theorem]{{\bf Corollary}}
\newtheorem{definition}[theorem]{{\bf Definition}}
\newtheorem{example}[theorem]{{\bf Example}}
\newtheorem{lemma}[theorem]{{\bf Lemma}}
\newtheorem{proposition}[theorem]{{\bf Proposition}}
\newtheorem{remark}[theorem]{{\bf Remark}}
\newtheorem{fact}[theorem]{Fact}

\fi

\newtheorem{fact}[theorem]{Fact}

\usepackage{stmaryrd,rotating,mathrsfs}

%\usepackage[active]{srcltx}

\iffalse
\newcommand{\problemx}[3]{
\par\noindent\underline{\sc#1}\par\nobreak\vskip.2\baselineskip
\begingroup\clubpenalty10000\widowpenalty10000
\setbox0\hbox{\bf INPUT: }\setbox1\hbox{\bf QUESTION: }
\dimen0=\wd0\ifnum\wd1>\dimen0\dimen0=\wd1\fi
\vskip-\parskip\noindent
\hbox to\dimen0{\box0\hfil}\hangindent\dimen0\hangafter1\ignorespaces#2\par
\vskip-\parskip\noindent
\hbox to\dimen0{\box1\hfil}\hangindent\dimen0\hangafter1\ignorespaces#3\par
\endgroup}

\newcommand{\problemy}[3]{
\par\noindent\underline{\sc#1}\par\nobreak\vskip.2\baselineskip
\begingroup\clubpenalty10000\widowpenalty10000
\setbox0\hbox{\bf INPUT: }\setbox1\hbox{\bf OUTPUT: }
\dimen0=\wd0\ifnum\wd1>\dimen0\dimen0=\wd1\fi
\vskip-\parskip\noindent
\hbox to\dimen0{\box0\hfil}\hangindent\dimen0\hangafter1\ignorespaces#2\par
\vskip-\parskip\noindent
\hbox to\dimen0{\box1\hfil}\hangindent\dimen0\hangafter1\ignorespaces#3\par
\endgroup}

\fi

\newcommand{\problemx}[3]{
\par\noindent\underline{\sc#1}\par\nobreak\vskip.2\baselineskip
\begingroup\clubpenalty10000\widowpenalty10000
\setbox0\hbox{\bf INPUT:\ }\setbox1\hbox{\bf QUESTION:\ }
\dimen0=\wd0\ifnum\wd1>\dimen0\dimen0=\wd1\fi
\vskip-\parskip\noindent
\hbox to\dimen0{\box0\hfil}\hangindent\dimen0\hangafter1\ignorespaces#2\par
\vskip-\parskip\noindent
\hbox to\dimen0{\box1\hfil}\hangindent\dimen0\hangafter1\ignorespaces#3\par
\endgroup}

\newcommand{\problemy}[3]{
\par\noindent\underline{\sc#1}\par\nobreak\vskip.2\baselineskip
\begingroup\clubpenalty10000\widowpenalty10000
\setbox0\hbox{\bf INPUT:\ }\setbox1\hbox{\bf OUTPUT:\ }
\dimen0=\wd0\ifnum\wd1>\dimen0\dimen0=\wd1\fi
\vskip-\parskip\noindent
\hbox to\dimen0{\box0\hfil}\hangindent\dimen0\hangafter1\ignorespaces#2\par
\vskip-\parskip\noindent
\hbox to\dimen0{\box1\hfil}\hangindent\dimen0\hangafter1\ignorespaces#3\par
\endgroup}

\renewcommand{\mod}{\text{mod }}
\newcommand{\modulo}{\mod}

%\renewcommand{\vec}{}

\def\AC{{\mathrm{AC}}}
\def\AG{{\mathsf{AG}}}
\def\AND{{\mathrm{AND}}}
\def\BIN{{\mathrm{BIN}}}
\def\CRR{{\mathrm{CRR}}}
\def\CTL{{\text{CTL}}}
\def\div{{\mathrm{div}}}
\def\dom{{\mathrm{dom}}}
\def\EF{{\mathsf{EF}}}
\def\AX{{\mathsf{AX}}}
\def\EXPTIME{{\mathsf{EXPTIME}}}
\def\inp{{\mathrm{in}}}
\def\leaf{{\mathrm{leaf}}}
\def\MAJ{{\mathrm{MAJ}}}
\def\NC{{\mathrm{NC}}}
\def\Omc{{\mathbb{O}}}
\def\OR{{\mathrm{OR}}}
\def\out{{\mathrm{out}}}
\def\Pmc{{\mathcal{P}}}
\def\PSPACE{{\mathsf{PSPACE}}}
\def\TC{{\mathrm{TC}}}
\def\U{{\mathsf{U}}}
\def\WU{{\mathsf{WU}}}
\def\X{{\mathsf{X}}}
\def\EX{{\mathsf{EX}}}
\def\E{{\mathsf{E}}}

\newcommand{\prop}{\rho}

\renewcommand{\L}{\mathsf{LOGSPACE}}

%% \usepackage[thmmarks]{ntheorem}
%% %%% Theorem environments
%% \newtheorem{lemma}{Lemma}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{proposition}{Proposition}
%% \newtheorem{corollary}{Corollary}

%% \theoremstyle{definition}
%% \theoremsymbol{\ensuremath{\lozenge}}
%% \theorembodyfont{\normalfont}
%% \newtheorem{definition}{Definition}

%% \theoremstyle{nonumverplain}
%% \theoremheaderfont{\normalfont\bf}
%% \theorembodyfont{\normalfont}
%% \theoremseparator{.}
%% \theoremsymbol{\ensuremath{_\blacksquare}}
%% \newtheorem{example}{Example}

%% \theoremstyle{nonumberplain}
%% \theoremheaderfont{\normalfont\itshape}
%% \theorembodyfont{\normalfont}
%% \theoremseparator{.}
%% \theoremsymbol{\ensuremath{_\Box}}
%% \newtheorem{proof}{Proof}

% Symbols
\newcommand{\naturals}{\ensuremath{\mathbf{N}}}
\newcommand{\integers}{\ensuremath{\mathbf{Z}}}

% Machines and their ingredients
\newcommand{\automaton}{\ensuremath{A}}
\newcommand{\clocations}{\ensuremath{Q}}
\newcommand{\iclocation}{\ensuremath{q_{\mathit{in}}}}
\newcommand{\fclocation}{\ensuremath{F}}
\newcommand{\transitionrel}{\ensuremath{\Delta}}
\newcommand{\transitionlabel}{\ensuremath{\lambda}}
\newcommand{\clocationlabel}{\ensuremath{\xi}}
\newcommand{\operations}{\ensuremath{Op}}
\newcommand{\operation}{\ensuremath{op}}
\newcommand{\add}[1]{\ensuremath{\mathsf{add}(#1)}}
\newcommand{\zero}{\ensuremath{\mathsf{zero}}}
\newcommand{\propositions}{\ensuremath{L}}
\newcommand{\configuration}{\ensuremath{c}}
\newcommand{\configurations}{\ensuremath{\mathcal{C}}}
\newcommand{\run}{\ensuremath{\pi}}
\newcommand{\trace}{\ensuremath{\tau}}
\newcommand{\pocatuple}{\ensuremath{\langle \clocations,
\iclocation, \fclocation, \transitionrel, \transitionlabel,
\clocationlabel \rangle}}

\newcommand{\circuit}{\ensuremath{\mathbb{C}}}

\newcommand{\tm}{\ensuremath{\mathcal{M}}}

%%% Names
\newcommand{\oca}{\textsc{OCA} }
\newcommand{\poca}{\textsc{POCA} }
\newcommand{\Poca}{\mathbb{P}}
\newcommand{\Soca}{\mathbb{S}}

%%% Problems
\newcommand{\stsat}{\textsc{Succinct 3-SAT} }
\newcommand{\tsat}{\textsc{3-SAT} }
\newcommand{\ltlpoca}{\textsc{LTL-POCA-MC} }

%%% Complexity classes
\newcommand{\nexptime}{\textsc{NExpTime} }
\newcommand{\conexptime}{\textsc{co-NExpTime} }
\newcommand{\pspace}{\textsc{PSpace} }

%%% LTL
\newcommand{\ltl}{\text{LTL}}
\newcommand{\ctl}{\text{CTL}}
%\newcommand{\until}{\ensuremath{\mathcal{~U~}}}
\newcommand{\until}{\ensuremath{\U}}
\newcommand{\eventually}{\ensuremath{\mathsf{F}}}
\newcommand{\globally}{\ensuremath{\mathsf{G}}}
\newcommand{\propalphabet}{\ensuremath{\Sigma}}
\newcommand{\pastarred}{\ensuremath{(\mathcal{P}(\propalphabet))^{\omega}}}

%%% Logic
\newcommand{\true}{\ensuremath{\mathtt{tt}}}
\newcommand{\false}{\ensuremath{\mathtt{ff}}}

%%% Control commands
\newcommand{\defemph}[1]{\emph{#1}}

%\newtheorem{fact}{Fact}

\newcommand{\LCM}{\text{LCM}}

\begin{document}

\title{Model Checking Succinct and Parametric One-Counter Automata}
\author{Stefan G\"oller\inst{1} \and
Christoph Haase\inst{2} \and
Jo\"el Ouaknine\inst{2} \and
James Worrell\inst{2}
}

\institute{
Universit\"at
Bremen, Institut f\"ur Informatik, Germany%\\
%\email{goeller@informatik.uni-bremen.de}
\and
Oxford University Computing Laboratory, UK
}

\maketitle

\begin{abstract}
  We investigate the decidability and complexity of various model
  checking problems over one-counter automata. More specifically, we
  consider \emph{succinct} one-counter automata, in which additive
  updates are encoded in binary, as well as \emph{parametric}
  one-counter automata, in which additive updates may be given as
  unspecified parameters. We fully determine the complexity of model
  checking these automata against CTL, LTL, and modal $\mu$-calculus
  specifications.
\end{abstract}

\section{Introduction}

@@

In Minsky's original formulation, counters were represented as integer
variables that could be incremented, decremented, or tested for
equality with zero by the finite-state controller. More recently,
driven by complexity-theoretic considerations on the one hand, and
potential applications on the other, researchers have investigated
additional primitive operations on counters, such as additive updates
encoded in binary~\cite{BHIMV06,LS08-atva} or even
in \emph{parametric} form, i.e., whose precise values depend on
parameters~\cite{BIL06-icalp,IJTW93-icalp}. We refer to such counter automata as
\emph{succinct} and \emph{parametric} respectively, the former being
viewed as a subclass of the latter. Natural applications of such
counter machines include the modeling of resource-bounded processes,
programs with lists, recursive or multi-threaded programs, and XML
query evaluation; see, e.g.,
\cite{ID06-tcs,BHIMV06,CR04}.

@@

The aim of the present paper is to study the decidability and
complexity of \emph{model checking} for succinct and parametric
one-counter automata. In view of Minsky's result, we restrict our
attention to \emph{succinct one-counter automata (SOCA)} and
\emph{parametric one-counter automata (POCA)}. On the specification
side, we focus on the three most prominent formalisms in the
literature, namely the temporal logics CTL and LTL, as well as the
modal $\mu$-calculus. For a counter automaton $\mathbb{A}$ and a
specification $\varphi$, we therefore consider the question of
deciding whether $\mathbb{A} \models \varphi$,
in case of POCA for all values of the parameters,
%\footnote{In the case of
%parametric automata, one asks whether the automaton satisfies the
%formula \emph{for all} legal values of the parameters.}
and investigate both the \emph{data} complexity (in which the formula
$\varphi$ is fixed) as well as the \emph{combined} complexity of this
problem. Our main results are summarized in Table \ref{T results}.
\begin{table}
\begin{center}
\begin{tabular}{ >{\centering\arraybackslash}m{3cm}
  >{\centering\arraybackslash}m{2cm} |
  >{\centering\arraybackslash}m{3.3cm} |
> {\centering\arraybackslash}m{3.3cm} |}\\
  &  &{\bf SOCA}&{\bf POCA}\\ \hline
{\raisebox{-2ex}[0ex]{{\bf CTL, $\mu$-calculus}}}&   {\bf data} &   & \\ \cline{2-2}
{\raisebox{-2ex}[-5ex]{{$\phantom{\text{CTL}}$}}}&  {\bf combined}&
{\raisebox{2ex}[0ex]{$\EXPSPACE$-complete}} &
{\raisebox{2ex}[0ex]{ $\Pi^0_1$-complete}}\\ \cline{1-4}
{\raisebox{-2ex}[0ex]{{\bf LTL}}}&  {\bf data} &
\multicolumn{2}{c|} {\raisebox{0ex}[0ex]{$\coNP$-complete}}
\\ \cline{2-4}
{\raisebox{-2ex}[-5ex]{{$\phantom{\text{CTL}}$}}}&  {\bf combined}&
$\PSPACE$-complete &
$\coNEXP$-complete\\ \cline{1-4}
\end{tabular}
\ \\[0.4cm]
\caption{The complexity of CTL, the modal $\mu$-calculus, and LTL on SOCA and
POCA.}{\label{T results}}
\end{center}
\end{table}

@@

We note some interesting differences between our results and
corresponding questions regarding finite automata. For the latter, the
(combined) model checking problems for CTL, the $\mu$-calculus, and
LTL are respectively known to be $\mathsf{P}$-complete, in
$\mathsf{NP} \cap \coNP$, and $\PSPACE$-complete. Somewhat
surprisingly, for SOCA and POCA, the complexity ordering is reversed
and LTL becomes easier to model check than either CTL or the
$\mu$-calculus.

On a technical level, the most intricate result is the
$\EXPSPACE$-hardness of CTL model checking for SOCA, which requires
several steps. We first show that $\mathsf{EXPSPACE}$ is `exponentially
$\mathsf{LOGSPACE}$-serializable', adapting the known proof that
$\PSPACE$ is $\mathsf{LOGSPACE}$-serializable. Unfortunately, and in
contrast to~\cite{GoLo10}, this does not immediately provide
an $\EXPSPACE$ lower bound. In a subsequent delicate stage of the
proof, we show how to `split' the counter in order simultaneously to
perform $\PSPACE$ computations in the counter and manipulate numbers
of exponential size in a SOCA of polynomial size.

Our paper is organized as follows. In Section \ref{S Prelim} we
introduce general notations.
Section \ref{S CTL} deals with CTL and the modal $\mu$-calculus.
LTL model checking is content of Section \ref{S LTL}.
Finally, in Section \ref{S Concl} we give conclusions.

\section{Preliminaries}{\label{S Prelim}}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\renewcommand{\O}{\mathbb{O}}
%\renewcommand{\P}{\mathcal{P}}

\newcommand{\trans}[1]{\stackrel{#1}{\longrightarrow}}
\newcommand{\hook}[1]{\stackrel{#1}{\hookrightarrow}}
\renewcommand{\rule}[2]{\stackrel{#2}{\stackrel{#1}{\longrightarrow}}}

\renewcommand{\M}{\mathcal{M}}

\newcommand{\bit}{\text{bit}}
\renewcommand{\succ}{\text{succ}}

\newcommand{\Reach}{\text{Reach}}
\newcommand{\Moves}{\text{Moves}}
\newcommand{\bin}{\text{bin}}

By $\Z$ we denote the {\em integers} and by $\N=\{0,1,2,\ldots\}$ the
denote the {\em naturals}.  For each $i,j\in\Z$ we define
$[i,j]=\{k\in\Z\mid i\leq k\leq j\}$ and $[j]=[1,j]$.  For each
$i,n\in\N$, let $\bit_i(n)$ denote the $i^{\text{th}}$ least
significant bit of the binary representation of $n$. Hence
$n=\sum_{i\in\N}2^i\cdot\bit_i(n)$.  By $\bin_m(n)=\bit_0(n)\cdots
\bit_{m-1}(n)$ we denote the first $m$ least significant bits written
from {\em left to right}. When $m$ is not important we just write
$\bin(n)$.
%For any non-empty subset $M\subseteq\Z^+$, define $\gcd(M)$ to be the largest
%$k\geq 1$ that divides every element of $M$.
%We define $\gcd(\emptyset)=0$.
Let $p_i$ denote the $i^{\text{th}}$ prime number for each $i\geq 1$, i.e. $p_1=2$,
$p_2=3$ and so on. We define $\log(n)=\min\{i\geq 1\mid 2^i> n\}$, in other
words $\log(n)$ denotes the number of bits that are needed to represent $n$ in
binary.
All polynomials $p:\N\rightarrow\N$ that occur in this paper are assumed to satisfy
$p(n)\geq n$ for each $n\in\N$.
For each word $v=a_1\cdots a_n\in\Sigma^n$ over some finite alphabet $\Sigma$
and each $i,j\in[n]$ define $v[i,j]=a_i\cdots a_j$ and
$v(i)=v[i,i]$.

\bigskip

\renewcommand{\l}{\triangleright}
\renewcommand{\r}{\triangleleft}
\newcommand{\move}{\vdash}
\noindent
{\bf Turing machines and complexity theory:} In the following, we
introduce $f(n)$ space-bounded deterministic Turing machines (DTMs) in
a suitable way for proving lower bounds. These contain precisely one
input tape and one working tape.  In our setting, the working alphabet is
 assumed to be $\{0,1,\l,\r\}$,
where $\l$ is the {\em left marker}
and $\r$ is the {\em right marker}.  The working tape of the initial
configuration of such a DTM on an input $w\in\Sigma^n$ is
assumed to be $\l0^{f(n)}\r$, whereas its input tape is $\l w\r$.
Before we define DTMs, let $\Upsilon=\{-1,0,+1\}$ denote a set of
directions.

Formally, a {\em $f(n)$-space bounded deterministic Turing machine (DTM)} is a tuple
$\M=(S,\Sigma,s_0,F,\mu)$, where $S$ is a finite set of {\em states},
$\Sigma$ is a {\em finite input alphabet} with $\l,\r\not\in\Sigma$,
$s_0\in S$ is an {\em initial state}, $F\subseteq S$ is a set of {\em final
states}, and with $\Sigma_\M=\Sigma\uplus\{\l,\r\}$ we have that
$$\mu\quad:\quad S\times\Sigma_\M\times\{0,1,\l,\r\}\ \rightarrow\
S\times\Upsilon^2\times\{0,1,\l,\r\}$$
is the {\em transition function}, where $\mu(s,b_1,b_2)=(s',\delta_1,\delta_2,b)$ means
that $\M$ is currently in state $s$, its input head reads $b_1$, its
working tape reads $b_2$, and $\M$ changes to state $s'$, moves its
input head in direction $\delta_1$,
moves its working head in direction $\delta_2$, and
writes the bit $b$.
%We require
%moreover that whenever $\M$'s input (resp. working) head reads $\l$, then $\M$
%must move its input (resp. working) head to the right, i.e. $+1$, and whenever
%$\M$'s input (resp.
%working) head reads $\r$, %it must hold that
%$\M$ moves its
%input (resp. working) head to the left, i.e. $-1$. Formally,
%whenever $\delta(s,b_1,b_2)=(s',\delta_1,\delta_2,b)$, then
%for each $i\in\{1,2\}$ we have that $b_i=\l$ implies $\delta_i=+1$ and $b_i=\r$
%implies $\delta_i=-1$.
Moreover, we require that $\M$ writes a marker precisely when it reads a marker,
more formally: $b_2=m$ if and only if $b=m$ for each $m\in\{\l,\r\}$.
As expected, a {\em configuration of $\M$} is a tuple $(s,\l w\r,\l w'\r,i,j)$,
where $s\in S$ is the current state,
$w\in\Sigma^n$ is an input,
$w'\in\{0,1\}^{f(n)}$ is the content of the working tape of $\M$,
$i\in[0,n+1]$ is the current input head position, and
$j\in[0,f(n)+1]$ is the current working head position.
%For configurations $C_1$ and $C_2$, we write $C_1\move_\M C_2$ whenever $\M$ can
%move from $C_1$ to $C_2$ on one step.
We define the {\em language $L(\M)$ of $\M$} to consist of all words
$w\in\Sigma^*$ such that
the {\em initial configuration}
$(s_0,\l w\r,\l 0^{f(|w|)}\r,0,0)$ reaches a configuration
$(s_0,\l w\r,\l w'\r,i,j)$ satisfying $s\in F$.

By $\L$, $\PSPACE$, and $\EXPSPACE$ we denote the class of all
problems that can be decided by DTM that
is logarithmically, polynomially, exponentially space bounded,
respectively. Recall that $\Sigma^0_1$ (resp. $\Pi^0_1$) is the class of
all languages that are (resp. whose complements are) recursively enumerable.
\newcommand{\Prop}{\mathcal{P}}
\newcommand{\Op}{\text{Op}}
%\newcommand{\add}{\mathsf{add}}
\renewcommand{\A}{\mathbb{A}}
%\newcommand{\EF}{\mathsf{EF}}
%\newcommand{\CTL}{\mathsh{CTL}}
\bigskip

\noindent
\textbf{Transition systems: }
In the following, we fix a countable set of atomic propositions $\Prop$.
A {\em transition system} is a tuple $T=(S,\{S_\prop\mid \prop\in\Prop\},\rightarrow)$, where
$S$ is a set of {\em states}, $S_\prop\subseteq S$ for each $\prop\in\Prop$ and
$S_\prop$ is
non-empty for finitely many $\prop\in\Prop$, and finally
$\rightarrow \subseteq S\times S$ is a set of {\em transitions}.
We prefer to use the infix notation $s_1\rightarrow s_2$ to abbreviate
$(s_1,s_2)\in\rightarrow$.
%A {\em finite path} of length $n$ is sequence of transitions
%$\pi=s_0\rightarrow s_1\cdots \rightarrow s_n$, where $n\in\N $.
 An {\em infinite path} is an infinite sequence
$\pi=s_0\rightarrow s_1\cdots$.
% We denote
%by $|\pi|$ the length of the path $\pi$, which we define to be $\omega$
%if $\pi$ is infinite.
For each infinite path
$\pi= s_0\rightarrow s_1 \rightarrow \cdots$ and
$i\in\N$, we denote by $\pi^i$ the path $s_i\rightarrow s_{i+1}\cdots$
and by $\pi(i)$ the state $s_i$.
Define the {\em trace of $\pi$} as $\trace(\pi):\N\rightarrow2^\Prop$ where
$\trace(i)=\{\prop\in\Prop\mid s_i\in S_\prop\}$ for each $i\in\N$.
\bigskip

\noindent
\textbf{Succinct and parametric one-counter automata: }
A {\em succinct one-counter automaton (SOCA)} is a tuple
 $\Soca=(Q,\{Q_\prop\mid \prop\in\Prop\},E,\lambda)$, where $Q$ is a finite set
of {\em control states}, $Q_\prop\subseteq Q$ for each $\prop\in\Prop$ and
$Q_\rho$ is non-empty for finitely many $\prop\in\Prop$,
$E\subseteq Q\times Q$ is a finite set of {\em transitions}, and
$\lambda:E\rightarrow \Z\cup\{\zero\}$ labels the edges with decrements,
increments, and zero tests.
A {\em parametric one-counter automaton (POCA)} is a tuple
$\Poca(X)=(Q,\{Q_\prop\mid\prop\in\Prop\},E,\lambda)$, where the first
three components are same as for a SOCA,
where $X$ is a finite set of {\em parameters over the natural numbers},
and where $\lambda:E\rightarrow
(\Z\cup\{\zero\}\cup\{\circ x\mid\circ\in\{+,-\}, x\in X\})$.
For each assignment $\sigma:X\rightarrow\N$  the
{\em induced SOCA} is defined as $\Poca^\sigma=(Q,\{Q_\prop\mid\prop\in\Prop\},E,\lambda')$
where $\lambda'(e)=\circ\sigma(x)$ whenever $\lambda(e)=\circ x$
for some $\circ\in\{+,-\}$ and $\lambda'(e)=\lambda(e)$ otherwise.
If $X=\{x\}$ is a singleton, then we also write $\Poca(x)$ for $\Poca(X)$.
The {\em size} of a POCA is defined as
$|\Poca|=|Q|+|X|+|E|\cdot\max\{\log(|a|)\mid a\in\lambda(E)\cap\Z\}$.
Hence, we represent each appearing
integer in binary.
The size of a SOCA is defined analogously.
A SOCA $\Soca=(Q,\{Q_\prop\mid \prop\in\Prop\},E,\lambda)$ describes a transition system $T(\Soca)=(Q\times\N,\{Q_\prop\times\N\mid
\prop\in\Prop\}, \rightarrow)$, where for each $q_1,q_2\in Q$ and each
$n_1,n_2\in\N$ we have $q_1(n_1)\rightarrow q_2(n_2)$ if and only
if either $\lambda(q_1,q_2)=n_2-n_1$, or
($n_1=n_2=0$ and $\lambda(q_1,q_2)=\zero)$.

\section{CTL Model Checking}{\label{S CTL}}

In Section \ref{S CTL syn} we introduce the syntax and semantics of $\ctl$.
Since the upper bounds for the modal $\mu$-calculus follow from
known results and since the lower bounds follow from CTL, we do not introduce
the modal $\mu$-calculus formally.
In Section \ref{S Upper} we state an $\EXPSPACE$ upper bound for the combined
complexity of $\ctl$ and the modal $\mu$-calculus on SOCA and a $\Pi^0_1$ upper bound
on POCA. In Section \ref{S Serial} we recall basic notions from complexity theory
such as serializability and results on Chinese remainder representation.
In Section \ref{S CTL hardness} we prove the main result of this section, namely
that the data complexity of $\ctl$ on SOCA is $\EXPSPACE$-hard.
Finally, we prove a $\Pi^0_1$ lower bound for the data complexity
of $\ctl$ on POCA in Section \ref{S CTL und}.

\subsection{$\ctl$: Syntax and Semantics}{\label{S CTL syn}}
\noindent
Formulas $\varphi$ of $\ctl$ are given by the following
grammar, where $\prop$ ranges over $\Prop$:

$$
\varphi\quad ::=\quad \prop\ \mid \neg\varphi\ \mid\ \varphi\wedge\varphi\ \mid\
\EX\varphi\ \mid \E (\varphi\U\varphi)\ \mid\
\E(\varphi\WU\varphi)
$$

\noindent
We have the following abbreviations:
$\varphi_1\vee\varphi_2=\neg(\neg\varphi_1\wedge\neg\varphi_2)$,
$\varphi_1\rightarrow\varphi_2=\neg\varphi_1\vee\varphi_2$,
$\true=\prop\vee\neg\prop$ for some atomic proposition $\prop\in\Prop$,
and $\EF\varphi=\E(\true\U\varphi)$.
The {\em size} of a $\ctl$ formula is inductively defined as follows:
$|\prop|=1$ for each $\prop\in\Prop$, $|\neg\varphi|=|\EX\varphi|=|\varphi|+1$,
$|\varphi_1\wedge\varphi_2|=|\E(\varphi_1\U\varphi_2)|=|\E(\varphi_1\WU\varphi_2)|=
|\varphi_1|+|\varphi_2|+1$.
Given a transition system $T=(S,\{S_\prop\mid \prop\in\Prop\},\rightarrow)$, a state
$s\in S$, and some $\CTL$ formula $\varphi$, define $(T,s)\models\varphi$ by
induction on the structure of $\varphi$ as follows:

\begin{eqnarray*}
  (T,s)\models \prop &  \quad\iff\quad & s\in S_p\quad \text{ for each
}\prop\in\Prop\\
  (T,s)\models \varphi_1\wedge\varphi_2 &  \quad\iff\quad &
(T,s)\models\varphi_1\text{ and }
(T,s)\models \varphi_2 \\
  (T,s)\models \neg\varphi &  \quad\iff\quad & (T,s)\not\models\varphi\\
  (T,s)\models \EX\varphi &  \quad\iff\quad & (T,t)\models\varphi
\text{ for some } t\in S \text{ with }s\rightarrow t\\
  (T,s)\models \E(\varphi_1\U\varphi_2) &  \quad\iff\quad &
\exists s_0,\cdots,s_n\in S, n\geq 0: s_0=s, (T,s_n)\models\varphi_2 \text{ such
that}\\
&& \forall i\in[0,n-1]: (T,s_i)\models\varphi_1\text{ and } s_i\rightarrow
s_{i+1}\\
  (T,s)\models \E(\varphi_1\WU\varphi_2) &  \quad\iff\quad &
(T,s)\models\E(\varphi_1\U\varphi_2)\ \text{ or }\\
&& \exists s_0,s_1,\ldots\in S:\forall i\geq 0: (T,s_{i})\models\varphi_1\text{ and }
s_i\rightarrow s_{i+1}
\end{eqnarray*}

\noindent
Let us define the {\em $\ctl$ model checking problem} on SOCA and POCA
respectively.\\

\problemx{$\ctl$ model checking on SOCA}
{SOCA $\Soca=(Q,\{Q_\prop\mid\prop\in\Prop\},E,\lambda)$, $q\in Q$, $n\in\N$ in binary, and a $\ctl$ formula
$\varphi$.}
{$(T(\Soca),q(n))\models\varphi$?}
\

\problemx{$\ctl$ model checking on POCA}
{POCA $\Poca(X)=(Q,\{Q_\prop\mid\prop\in\Prop\},E,\lambda)$, $q\in Q$, $n\in\N$ in binary, and a $\ctl$ formula
$\varphi$.}
{$(T(\Poca^\sigma),q(n))\models\varphi$ for every $\sigma:X\rightarrow\N$?}

\subsection{Upper bounds}{\label{S Upper}}
Let us first state upper bounds on model
checking CTL and the modal $\mu$-calculus on SOCA.
 {\em One-counter automata (OCA)} are SOCA in which
 the numbers that occur in the transition labels are represented  in unary
 instead in binary.
%In this section we state the upper bounds on model checking $\ctl$ on SOCA and
%POCA. They follow from known results or are immediate.
The following theorem gives an upper bound on model checking the
{\em modal $\mu$-calculus} on OCA.
Consult \cite{ArnNiw01} for more details on the modal $\mu$-calculus.

\begin{theorem}[\cite{Serr06}]
The combined complexity of the modal $\mu$-calculus on OCA is in $\PSPACE$.
\end{theorem}

\noindent
Since every SOCA can be transformed into an exponentially larger OCA
 and since each $\ctl$ formula can be translated into an alternation-free
$\mu$-calculus formula with a linear blowup, the following corollary is
immediate.

\begin{corollary}{\label{C CTL up}}
The combined complexity of $\ctl$ and the modal $\mu$-calculus on SOCA is in $\EXPSPACE$.
\end{corollary}

\noindent
The following upper bound for $\ctl$ on POCA is straightforward.

\begin{corollary}
The combined complexity of $\ctl$ and the modal $\mu$-calculus on POCA is in $\Pi^0_1$.
\end{corollary}
\begin{proof}
Let $\Poca(X)=(Q,\{Q_\prop\mid\prop\in\Prop\},E,\lambda)$ be a POCA,
 $q\in Q$, and $n\in\N$, and $\varphi$ be
a formula.
From Corollary \ref{C CTL up} it follows that
for each given $\sigma:X\rightarrow\N$ the question
$
(T(\Poca^\sigma),q(n))\models\varphi
$
is a decidable predicate.
Hence, the question if
$
(T(\Poca^\sigma),q(n))\models\varphi
$
for each $\sigma:X\rightarrow\N$ is a $\Pi^0_1$-predicate since
we can encode the set of all such assignments into the naturals
via G\"odel encoding.
\qed
\end{proof}

>>>>

@@

In Minsky's original formulation, counters were represented as integer
variables that could be incremented, decremented, or tested for
equality with zero by the finite-state controller. More recently,
driven by complexity-theoretic considerations on the one hand, and
potential applications on the other, researchers have investigated
additional primitive operations on counters, such as additive updates
encoded in binary~\cite{BHIMV06,LS08-atva} or even
in \emph{parametric} form, i.e., whose precise values depend on
parameters~\cite{BIL06-icalp,IJTW93-icalp}. We refer to such counter automata as
\emph{succinct} and \emph{parametric} resp., the former being
viewed as a subclass of the latter. Natural applications of such
counter machines include the modelling of resource-bounded processes,
programs with lists, recursive or multi-threaded programs, and XML
query evaluation; see, e.g.,
\cite{BHIMV06,CR04,ID06-tcs}.

@@

The aim of the present paper is to study the decidability and
complexity of \emph{model checking} for succinct and parametric
one-counter automata. In view of Minsky's result, we restrict our
attention to \emph{succinct one-counter automata (SOCA)} and
\emph{parametric one-counter automata (POCA)}. On the specification
side, we focus on the three most prominent formalisms in the
literature, namely the temporal logics CTL and LTL, as well as the
modal $\mu$-calculus. For a counter automaton $\mathbb{A}$ and a
specification $\varphi$, we therefore consider the question of
deciding whether $\mathbb{A} \models \varphi$,
in case of POCA for all values of the parameters,
%\footnote{In the case of
%parametric automata, one asks whether the automaton satisfies the
%formula \emph{for all} legal values of the parameters.}
and investigate both the \emph{data} complexity (in which the formula
$\varphi$ is fixed) as well as the \emph{combined} complexity of this
problem. Our main results are summarized in Table \ref{tab:results}.
\begin{table}[t]
\begin{center}
\begin{tabular}{ >{\centering\arraybackslash}m{3cm}
  >{\centering\arraybackslash}m{2cm} |
  >{\centering\arraybackslash}m{3.3cm} |
> {\centering\arraybackslash}m{3.3cm} |}\\
  &  &{\bf SOCA}&{\bf POCA}\\ \hline
{\raisebox{-2ex}[0ex]{{\bf CTL, $\mu$-calculus}}}&   {\bf data} &   & \\ \cline{2-2}
{\raisebox{-2ex}[-5ex]{{$\phantom{\text{CTL}}$}}}&  {\bf combined}&
{\raisebox{2ex}[0ex]{$\EXPSPACE$-complete}} &
{\raisebox{2ex}[0ex]{ $\Pi^0_1$-complete}}\\ \cline{1-4}
{\raisebox{-2ex}[0ex]{{\bf LTL}}}&  {\bf data} &
\multicolumn{2}{c|} {\raisebox{0ex}[0ex]{$\coNP$-complete}}
\\ \cline{2-4}
{\raisebox{-2ex}[-5ex]{{$\phantom{\text{CTL}}$}}}&  {\bf combined}&
$\PSPACE$-complete &
$\coNEXP$-complete\\ \cline{1-4}
\end{tabular}
\end{center}
\label{tab:results}
\caption{The complexity of CTL, the modal $\mu$-calculus, and LTL
             on SOCA and POCA.}
\end{table}

@@

%% We note some interesting differences between our results and
%% corresponding questions regarding finite automata. For the latter, the
%% (combined) model checking problems for CTL, the $\mu$-calculus, and
%% LTL are respectively known to be $\mathsf{P}$-complete, in
%% $\mathsf{NP} \cap \coNP$, and $\PSPACE$-complete. Somewhat
%% surprisingly, for SOCA and POCA, the complexity ordering is reversed
%% and LTL becomes easier to model check than either CTL or the
%% $\mu$-calculus.

On a technical level, the most intricate result is the
$\EXPSPACE$-hardness of CTL model checking for SOCA, which requires
several steps. We first show that $\mathsf{EXPSPACE}$ is `exponentially
$\mathsf{LOGSPACE}$-serializable', adapting the known proof that
$\PSPACE$ is $\mathsf{LOGSPACE}$-serializable. Unfortunately, and in
contrast to~\cite{GoLo10}, this does not immediately provide
an $\EXPSPACE$ lower bound. In a subsequent delicate stage of the
proof, we show how to partition the counter in order simultaneously to
perform $\PSPACE$ computations in the counter and manipulate numbers
of exponential size in a SOCA of polynomial size.

%For reasons of space, we have had to abbreviate or omit a number of
%proofs; full details can however be found in the technical
%report~\cite{techrep}.
